Query Optimization for Dynamic Imputation

### 2017-05-11 paper/experiments.tex
sprinkled remaining todos, to make sure we don't miss

For each query, we evaluate three different configurations: \ProjectName{} optimizing for
quality ($\alpha=0$), \ProjectName{} optimizing for performance ($\alpha=1$), and imputing
at the base tables followed by executing the query traditionally.

\Cref{fig:runtimes} shows a summary of the performance results. The quality-optimized
queries are an order-of-magnitude faster than imputing on the base tables. We can get
another order-of-magnitude speedup when optimizing for performance. This performance
differential means it is feasible
to explore multiple imputations operations (including more expensive operators) when using
\ProjectName{}, in contrast to the traditional approach of base table imputation.

\begin{figure}
\includegraphics[width=\columnwidth]{figures/running_times_combined_bar.png}
\caption{Query runtimes with imputation on base tables, \ProjectName{}
    optimizing for quality ($\alpha=0$), and \ProjectName{} optimizing for
    performance ($\alpha=1$). Quality-optimized, on-the-fly imputation provides an order of
    magnitude speedup over imputing on base tables. Efficiency-optimized on-the-fly
    imputation can provide another order-of-magnitude speedup at the expense of quality.}

\label{fig:runtimes}
\end{figure}

@@

In the median across a variety of queries and choices of $\alpha$, planning took
\planningmediantime{} ms, 
constituting \planningruntimepercent{} percent of total runtime. In all cases, the optimizer
returned a query plan within \planningmaxtime{} ms, with times similar across values of $\alpha$.

@@

For $\alpha=0$ and $\alpha=1$, we pair up tuples and compute SMAPE for all queries. That is,
we compute tuple-wise absolute percentage deviations within each iteration of a query, and
we report this value averaged over all iterations.  We
can see that optimizing for quality indeed reduces the SMAPE of query results.  In general, the SMAPE relative 
to the base-imputation approach is low---between \lowsmapealphazero{} and
\highsmapealphaoneexacs{} percent---indicating that on-the-fly imputation produces similar
results to imputation at the base tables. Notably, the results with largest error correspond
most closely to one existing approach of dropping all \nullv{}s.

@@

\begin{table}
\centering
\input{tables/perf_summary.tex}
\caption{Symmetric-Mean-Absolute-Percentage-Error for queries run under different $\alpha$
    parameterizations. To calculate SMAPE, \ProjectName{} results are compared to query results returned
    from executions with base table imputation. Queries optimized
    for quality ($\alpha=0$) generally achieve lower error than queries optimized for
    efficiency ($\alpha=1$). The count fraction column shows the number of tuples used in calculating each aggregate
     as a fraction of the number of tuples used when running the same query after imputing on the base table.
    A lower count share reflects more potential for errors.}
\label{table:smape}
\end{table}

@@

\subsubsection{Planning times}
\begin{table}
\centering
\input{tables/planning_summary_exact_pareto.tex}
\caption{Planning times are sub-second for up to 4 joins. For 5+  joins planning times rise. Many practical queries
can be formulated with a smaller number of joins, but to support more complex queries \ProjectName{} provides an
optimization that maintains approximate Pareto sets for queries with many joins. This reduces the 29 second
planning time shown for 7 joins by approximately 45\%, on average.}
\label{table:planning-times}
\end{table}

We explored the practical implications of our planning algorithm by constructing a series of queries with increasing number of joins.
We used the CDC tables and evaluated a combination of self-joins and joins across reference tables. Planning times for queries
involving 1 to 5 joins, which although limited is a practical value for real world exploratory queries, take less than a second.
For queries with 7 joins, planning time increases substantially to 29 seconds, on average. We introduced an optimization that reduces
the plan space by maintaining only approximations of the Pareto set for queries involving a large number of joins.
With this optimization, the same set of queries have an average planning time of 16 seconds (2 seconds SE).

>>>>

For each query, we evaluate \todo{update: three different configurations}: \ProjectName{} optimizing for
quality ($\alpha=0$), \ProjectName{} optimizing for performance ($\alpha=1$), and imputing
at the base tables followed by executing the query traditionally.

\Cref{fig:runtimes} shows a summary of the performance results. The quality-optimized
queries are an \todo{update/validate: order-of-magnitude faster than imputing on the base tables. We can get
another order-of-magnitude speedup when optimizing for performance}. This performance
differential means it is feasible
to explore multiple imputations operations (including more expensive operators) when using
\ProjectName{}, in contrast to the traditional approach of base table imputation.

\begin{figure}
\includegraphics[width=\columnwidth]{figures/running_times_combined_bar.png}
\caption{Query runtimes with imputation on base tables, \ProjectName{}
    optimizing for quality ($\alpha=0$), and \ProjectName{} optimizing for
    performance ($\alpha=1$). Quality-optimized, on-the-fly imputation provides an order of
    magnitude speedup over imputing on base tables. Efficiency-optimized on-the-fly
    imputation can provide another order-of-magnitude speedup at the expense of quality.}
\todo{update with new figure}   
\label{fig:runtimes}
\end{figure}

@@

\todo{update: In the median across a variety of queries and choices of $\alpha$, planning took
\planningmediantime{} ms, 
constituting \planningruntimepercent{} percent of total runtime. In all cases, the optimizer
returned a query plan within \planningmaxtime{} ms, with times similar across values of $\alpha$.}

@@

\todo{Distinguish between underlying imputation methods}
For $\alpha=0$ and $\alpha=1$, we pair up tuples and compute SMAPE for all queries. That is,
we compute tuple-wise absolute percentage deviations within each iteration of a query, and
we report this value averaged over all iterations.  We
can see that optimizing for quality indeed reduces the SMAPE of query results.  In general, the SMAPE relative 
to the base-imputation approach is low---\todo{update: between \lowsmapealphazero{} and
\highsmapealphaoneexacs{} percent}---indicating that on-the-fly imputation produces similar
results to imputation at the base tables. Notably, the results with largest error correspond
most closely to one existing approach of dropping all \nullv{}s.

@@

\begin{table}
\centering
\input{tables/perf_summary.tex}
\caption{Symmetric-Mean-Absolute-Percentage-Error for queries run under different $\alpha$
    parameterizations. To calculate SMAPE, \ProjectName{} results are compared to query results returned
    from executions with base table imputation. Queries optimized
    for quality ($\alpha=0$) generally achieve lower error than queries optimized for
    efficiency ($\alpha=1$). The count fraction column shows the number of tuples used in calculating each aggregate
     as a fraction of the number of tuples used when running the same query after imputing on the base table.
    A lower count share reflects more potential for errors.}
\todo{update with new figure}
\label{table:smape}
\end{table}

@@

\subsubsection{Planning times}
\begin{table}
\centering
\input{tables/planning_summary_exact_pareto.tex}
\caption{Planning times are sub-second for up to 4 joins. For 5+  joins planning times rise. Many practical queries
can be formulated with a smaller number of joins, but to support more complex queries \ProjectName{} provides an
optimization that maintains approximate Pareto sets for queries with many joins. This reduces the 29 second
planning time shown for 7 joins by approximately 45\%, on average.
\todo{update with new table and values}
}
\label{table:planning-times}
\end{table}

We explored the practical implications of our planning algorithm by constructing a series of queries with increasing number of joins.
We used the CDC tables and evaluated a combination of self-joins and joins across reference tables. Planning times for queries
involving 1 to 5 joins, which although limited is a practical value for real world exploratory queries, take less than a second.
\todo{update: For queries with 7 joins, planning time increases substantially to 29 seconds, on average. We introduced an optimization that reduces
the plan space by maintaining only approximations of the Pareto set for queries involving a large number of joins.
With this optimization, the same set of queries have an average planning time of 16 seconds (2 seconds SE).}
